\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-24 doc/incertae_sedis.tex
Some rearranging, updating and more reformatting.

The rooted split derived from a node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for $n$ on the resulting synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.

\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}

@@

\paragraph{Information content of the \emph{incertae sedis} annotation}

@@

\paragraph{Naive interpretation: ignoring the annotation}

\mthcomment{ Do we have some numbers on this (presumably from just
running the only propinquity with a different set of prune flags)? }

\paragraph{Rich taxon-specific attachment points}

As mentioned above, the \emph{incertae sedis} annotation may mean that a taxon is
unplaced, but only a few placements within the taxonomy are possible.
For example, a genus might be characterized as \emph{incertae sedis}
at the level of the subfamily.
Thus, it could be placed inside any of the
hypothesized sub-families, or within its own subfamily, but it would
be unexpected for it to be placed inside one of the other genera.

@@

@@

\subsection{Placement causes broken taxa}

Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies.
Since
phylogenies are ranked higher than taxonomy, they may place members of
incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa.
However, without the
incertae sedis semantics described above, placing a taxon $A$ within a
taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has
three main effects: \begin{enumerate}
    \item The edge leading to clade $B$ is lost.
    \item Any taxonomy-only children of $B$ are placed at the
        MRCA of the phylogenetically informed taxa, instead of at node $B$.
    \item The name for taxon $B$ is lost.
    \item We do not record $A$ as being placed within $B$.
\end{enumerate} In order to avoid a situation
where input phylogenies conflict with a large number of taxa when
\emph{incertae sedis} taxa are placed within them, we have previously
filtered \emph{incertae sedis} taxa from the taxonomy when
constructing all prior synthesis trees.

>>>>

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.

\paragraph{(Not using rank information) Rich taxon-specific attachment points}
% Question: why not use rank information?  Answer: well, we could, but lets start simple.
As mentioned above, the \emph{incertae sedis} annotation may mean that a taxon is
unplaced, but only a few placements within the taxonomy are possible.
For example, a genus might be characterized as \emph{incertae sedis}
at the level of the subfamily.
Thus, it could be placed inside any of the
hypothesized sub-families, or within its own subfamily, but it would
be unexpected for it to be placed inside one of the other genera.

\paragraph{(Split-based semantics) Equivalence of a taxonomy tree and a list of rooted splits}

@@

\paragraph{(A first semantics) Naive interpretation: ignoring the annotation}

% \mthcomment{ Do we have some numbers on this (presumably from just
%running the only propinquity with a different set of prune flags)? }

% \bdrcomment{ I think we would currently lose about 10 taxa }

\bdrcomment{The reason why we don't do the naive thing is currently in section \ref{conflict}}
\paragraph{(Split-based semantics) Information content of the \emph{incertae sedis} annotation}

@@

@@

\subsection{Surprising results of the split-based semantics}

1. Inter-digitation

2. Placing A next to B, then B within A.

@@

\subsection{Placement causes broken taxa}
\label{conflict}
Synthesis with \emph{incertae sedis} taxa has the potential to resolve
uncertain taxon placements using information from phylogenies.
Since
phylogenies are ranked higher than taxonomy, they may place members of
incertae sedis clades as they wish even when we do not take into
account the semantics of incertae sedis taxa.
However, without the
incertae sedis semantics described above, placing a taxon $A$ within a
taxon $B$ results in conflict with taxon $B$ in the taxonomy.
This has
three main effects: \begin{enumerate}
    \item The edge leading to clade $B$ is lost.
    \item Any taxonomy-only children of $B$ are placed at the
        MRCA of the phylogenetically informed taxa, instead of at node $B$.
    \item The name for taxon $B$ is lost.
    \item We do not record $A$ as being placed within $B$.
\end{enumerate} In order to avoid a situation
where input phylogenies conflict with a large number of taxa when
\emph{incertae sedis} taxa are placed within them, we have previously
filtered \emph{incertae sedis} taxa from the taxonomy when
constructing all prior synthesis trees.
